---
title: "BASS-BO vs GP-BO: One-Dimensional Comparison"
author: "Adrian TJ"
date: "June 2026"
output: 
  html_document:
    toc: true
    theme: united
---

# Introduction

This notebook compares **Bayesian Adaptive Spline Surfaces (BASS)** and **Gaussian Processes (GP)** as surrogate models for Bayesian Optimization in a one-dimensional setting. We evaluate their ability to navigate a complex, non-linear objective function with oscillations, a jump discontinuity, and local features.

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lhs)
library(BASS)
library(GPfit)
library(gt)
```

# Objective Function

The objective function $f(x)$ is defined on the unit interval $[0, 1]$. It includes:
- A sinusoidal base.
- A jump discontinuity at $x = 0.62$.
- A local bump and a local valley to test surrogate sensitivity.

```{r objective-and-settings}
set.seed(123)

#' Complex 1D Objective Function
f <- function(x) {
  x <- pmin(pmax(x, 0), 1)  # clamp to [0,1]
  term1 <- sin(10 * pi * x) * (1 - x)
  term2 <- 0.25 * cos(24 * pi * x^2)
  jump  <- ifelse(x > 0.62, 0.8, 0.0)
  bump  <- 1.2 * exp(-120 * (x - 0.18)^2)
  valley <- -1.0 * exp(-80 * (x - 0.78)^2)
  term1 + term2 + bump + valley + jump
}

# Experiment settings
n0 <- 5             # Initial design size
budget <- 80        # Number of sequential BO steps
grid_n <- 1000      # Resolution of the candidate grid
kappa <- 2.0        # LCB exploration parameter
eps <- 1e-8         # Small nugget for GP stability

x_grid <- seq(0, 1, length.out = grid_n)

# Helper: Find the index of the closest value on the grid
closest_idx <- function(x, grid) which.min(abs(grid - x))
```

# Initialization

Both methods start with the same initial design points generated via **Maximin Latin Hypercube Sampling (LHS)**.

```{r initialization}
x_init <- maximinLHS(n0, 1)[, 1]
y_init <- f(x_init)

init_df <- tibble(x = x_init, y = y_init)
init_df %>% gt() %>% tab_header("Initial Design Points")
```

# Optimization Algorithms

## BASS-BO
BASS uses Bayesian Adaptive Splines. It is flexible and handles discontinuities better than standard GPs because it doesn't assume a stationary kernel.

```{r bass-bo-function}
run_bass_bo <- function(x_init, y_init, x_grid, budget, kappa) {
  evals <- tibble(x = x_init, y = y_init)
  
  # Tracking grid usage to avoid re-evaluating same points
  used <- rep(FALSE, length(x_grid))
  for (x0 in evals$x) used[closest_idx(x0, x_grid)] <- TRUE
  
  best_so_far <- numeric(budget + 1)
  best_so_far[1] <- min(evals$y)
  
  for (t in 1:budget) {
    # Fit BASS (Degree 2 for interactions/curves)
    fit <- bass(matrix(evals$x, ncol = 1), evals$y, degree = 2, verbose = FALSE)
    
    # Predict on grid (posterior draws)
    pred <- predict(fit, newdata = data.frame(x = x_grid))
    pred_mat <- as.matrix(pred)
    if (ncol(pred_mat) != length(x_grid) && nrow(pred_mat) == length(x_grid)) pred_mat <- t(pred_mat)
    
    mu <- colMeans(pred_mat)
    sd_post <- apply(pred_mat, 2, sd)
    
    # Lower Confidence Bound
    lcb <- mu - kappa * sd_post
    lcb[used] <- Inf
    
    idx_next <- which.min(lcb)
    x_next <- x_grid[idx_next]
    y_next <- f(x_next)
    
    evals <- bind_rows(evals, tibble(x = x_next, y = y_next))
    used[idx_next] <- TRUE
    best_so_far[t + 1] <- min(evals$y)
  }
  list(evals = evals, best_so_far = best_so_far)
}
```

## GP-BO
Standard Gaussian Process regression with a Matérn-like kernel (implied by `GPfit`).

```{r gp-bo-function}
run_gp_bo <- function(x_init, y_init, x_grid, budget, kappa, eps = 1e-8) {
  evals <- tibble(x = x_init, y = y_init)
  used <- rep(FALSE, length(x_grid))
  for (x0 in evals$x) used[closest_idx(x0, x_grid)] <- TRUE
  
  best_so_far <- numeric(budget + 1)
  best_so_far[1] <- min(evals$y)
  
  for (t in 1:budget) {
    fit <- GP_fit(X = matrix(evals$x, ncol = 1), Y = evals$y)
    pred <- predict(fit, xnew = matrix(x_grid, ncol = 1))
    
    mu <- as.numeric(pred$Y_hat)
    sd_gp <- sqrt(pmax(as.numeric(pred$MSE), 0) + eps)
    
    lcb <- mu - kappa * sd_gp
    lcb[used] <- Inf
    
    idx_next <- which.min(lcb)
    x_next <- x_grid[idx_next]
    y_next <- f(x_next)
    
    evals <- bind_rows(evals, tibble(x = x_next, y = y_next))
    used[idx_next] <- TRUE
    best_so_far[t + 1] <- min(evals$y)
  }
  list(evals = evals, best_so_far = best_so_far)
}
```

## Random Search
Baseline for comparison.

```{r random-search-function}
run_random_search <- function(x_init, y_init, budget) {
  evals <- tibble(x = x_init, y = y_init)
  best_so_far <- numeric(budget + 1)
  best_so_far[1] <- min(evals$y)
  
  for (t in 1:budget) {
    x_next <- runif(1, 0, 1)
    y_next <- f(x_next)
    evals <- bind_rows(evals, tibble(x = x_next, y = y_next))
    best_so_far[t + 1] <- min(evals$y)
  }
  list(evals = evals, best_so_far = best_so_far)
}
```

# Execution

```{r experiment-run}
res_bass <- run_bass_bo(x_init, y_init, x_grid, budget, kappa)
res_gp   <- run_gp_bo(x_init, y_init, x_grid, budget, kappa, eps)
res_rand <- run_random_search(x_init, y_init, budget)
```

# Results

## Convergence Comparison

```{r convergence-plot, fig.width=10, fig.height=5}
conv_df <- bind_rows(
  tibble(iter = 0:budget, best = res_bass$best_so_far, method = "BASS-BO"),
  tibble(iter = 0:budget, best = res_gp$best_so_far,   method = "GP-BO"),
  tibble(iter = 0:budget, best = res_rand$best_so_far, method = "Random Search")
)

ggplot(conv_df, aes(x = iter, y = best, color = method)) +
  geom_line(linewidth = 1) +
  labs(title = "Convergence: Best Objective Found vs Iteration",
       x = "Iteration", y = "Min Objective") +
  theme_minimal()
```

## Posterior Visualizations

We visualize the final surrogate fit and uncertainty bands for each method.

```{r plotting-helpers, echo=FALSE}
# Internal helpers for generating plotting data
get_bass_posterior <- function(evals, x_grid, degree = 2) {
  fit <- bass(matrix(evals$x, ncol = 1), evals$y, degree = degree, verbose = FALSE)
  pred <- predict(fit, newdata = data.frame(x = x_grid))
  pred_mat <- as.matrix(pred)
  if (ncol(pred_mat) != length(x_grid) && nrow(pred_mat) == length(x_grid)) pred_mat <- t(pred_mat)
  list(mu = colMeans(pred_mat), sigma = apply(pred_mat, 2, sd))
}

get_gp_posterior <- function(evals, x_grid, eps = 1e-8) {
  fit <- GP_fit(X = matrix(evals$x, ncol = 1), Y = evals$y)
  pred <- predict(fit, xnew = matrix(x_grid, ncol = 1))
  list(mu = as.numeric(pred$Y_hat), sigma = sqrt(pmax(as.numeric(pred$MSE), 0) + eps))
}

plot_posterior_panel <- function(x_grid, evals, title, n0, mu = NULL, sigma = NULL, show_posterior = TRUE) {
  y_true <- f(x_grid)
  y_lim <- range(c(y_true, evals$y, if (show_posterior) c(mu+sigma, mu-sigma) else NA), na.rm = TRUE)
  
  plot(x_grid, y_true, type = "l", lwd = 1.5, xlab = "x", ylab = "f(x)", main = title, ylim = y_lim)
  if (show_posterior) {
    polygon(c(x_grid, rev(x_grid)), c(mu + sigma, rev(mu - sigma)), col = rgb(0, 0, 1, 0.20), border = NA)
    lines(x_grid, mu, lwd = 2, lty = "dotted")
  }
  
  # Color points by time (Red -> Green)
  cols <- grDevices::colorRampPalette(c("red", "green"))(nrow(evals))
  points(evals$x, evals$y, pch = 16, cex = 0.85, col = cols)
  points(evals$x[1:n0], evals$y[1:n0], pch = 1, cex = 1.05, lwd = 1) # Circle initial points
}
```

```{r final-visuals, fig.width=16, fig.height=5}
post_bass <- get_bass_posterior(res_bass$evals, x_grid)
post_gp   <- get_gp_posterior(res_gp$evals, x_grid)

par(mfrow = c(1, 3))
plot_posterior_panel(x_grid, res_bass$evals, "BASS-BO Fit", n0, post_bass$mu, post_bass$sigma)
plot_posterior_panel(x_grid, res_gp$evals, "GP-BO Fit", n0, post_gp$mu, post_gp$sigma)
plot_posterior_panel(x_grid, res_rand$evals, "Random Search (Evaluations)", n0, show_posterior = FALSE)
```















