---
title: "Multi-Dimensional Bayesian Optimization: BASS vs GP"
author: "Adrian TJ"
date: "June 2026"
output: 
  html_document:
    toc: true
    theme: readable
---

# Introduction

This notebook extends the comparison between **BASS-BO** and **GP-BO** to multi-dimensional spaces. We evaluate how the flexible basis functions of BASS compare against the kernel-based approach of Gaussian Processes in $d=3$ dimensions.

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lhs)
library(BASS)
library(GPfit)
library(gt)
```

# Objective Function

We define a $d$-dimensional synthetic function characterized by:
- A quadratic base "bowl".
- Multi-dimensional oscillations and interactions.
- Localized features (bumps/valleys) that require precise sampling.
- A jump discontinuity in the first dimension.

```{r objective-and-settings}
set.seed(1244)

# Dimension of the input space
d <- 3

#' Synthetic Multi-Dimensional Objective Function
f <- function(X) {
  X <- as.matrix(X)
  if (is.null(nrow(X))) X <- matrix(X, nrow = 1)
  
  # Quadratic trend
  base <- rowSums((X - 0.35)^2)
  
  # Nonlinear components
  osc1 <- sin(6 * pi * X[, 1]) * exp(-2 * X[, 1])
  osc2 <- if (ncol(X) >= 2) 0.35 * cos(5 * pi * X[, 2]^2) else 0
  inter <- if (ncol(X) >= 2) 0.5 * X[, 1] * X[, 2] else 0
  
  # Localized bumps and valleys
  c1 <- rep(0.2, ncol(X))
  c2 <- rep(0.75, ncol(X))
  bump   <- 1.2 * exp(-35 * rowSums((X - matrix(c1, nrow(X), ncol(X), byrow = TRUE))^2))
  valley <- -1.0 * exp(-28 * rowSums((X - matrix(c2, nrow(X), ncol(X), byrow = TRUE))^2))
  
  # Discontinuity
  jump <- ifelse(X[, 1] > 0.62, 0.4, 0)
  
  base + osc1 + osc2 + inter + bump + valley + jump
}

# Experiment Configuration
n0 <- max(2 * d + 1, 8)   # Size of initial space-filling design
budget <- 80              # Number of sequential BO steps
n_cand <- 1000            # Size of candidate pool for acquisition
kappa <- 2.0              # LCB exploration parameter
eps <- 1e-10              # Nugget for GP stability
dup_tol <- 1e-10          # Tolerance for duplicate detection

#' Utility: Check for points already evaluated
is_duplicate <- function(x, X, tol = 1e-10) {
  X <- as.matrix(X)
  xmat <- matrix(x, nrow = nrow(X), ncol = ncol(X), byrow = TRUE)
  any(rowSums((X - xmat)^2) <= tol^2)
}
```

# Initialization

```{r initialization}
# Generate initial design via Maximin LHS
X_init <- maximinLHS(n0, d)
y_init <- f(X_init)

init_df <- as_tibble(X_init, .name_repair = ~ paste0("x", seq_along(.))) %>%
  mutate(y = y_init)

init_df %>% gt() %>% tab_header(sprintf("Initial %d-D Design", d))
```

# Optimization Algorithms

## BASS-BO
BASS fits an additive or interaction model using splines, which can adapt to local features without a global stationary assumption.

```{r bass-bo-function}
run_bass_bo <- function(X_init, y_init, budget, n_cand, kappa, dup_tol = 1e-10) {
  X_eval <- as.matrix(X_init); y_eval <- as.numeric(y_init); d <- ncol(X_eval)
  best_so_far <- numeric(budget + 1); best_so_far[1] <- min(y_eval)
  
  for (t in 1:budget) {
    # Fit BASS surrogate
    fit <- bass(xx = X_eval, y = y_eval, degree = 2, verbose = FALSE)
    
    # Generate and predict on candidates
    X_cand <- maximinLHS(n_cand, d)
    pred <- predict(fit, newdata = as.data.frame(X_cand))
    pred_mat <- as.matrix(pred)
    if (ncol(pred_mat) != nrow(X_cand)) pred_mat <- t(pred_mat)
    
    # Acquisition: Lower Confidence Bound
    mu <- colMeans(pred_mat)
    sd_post <- apply(pred_mat, 2, sd)
    lcb <- mu - kappa * sd_post
    
    # Selection of best non-duplicate
    ord <- order(lcb)
    pick <- NA_integer_
    for (idx in ord) {
      if (!is_duplicate(X_cand[idx, ], X_eval, tol = dup_tol)) {
        pick <- idx; break
      }
    }
    if (is.na(pick)) pick <- ord[1]
    
    # Evaluate and Update
    x_next <- X_cand[pick, , drop = FALSE]; y_next <- f(x_next)
    X_eval <- rbind(X_eval, x_next); y_eval <- c(y_eval, y_next)
    best_so_far[t + 1] <- min(y_eval)
  }
  list(X_eval = X_eval, y_eval = y_eval, best_so_far = best_so_far)
}
```

## GP-BO

```{r gp-bo-function}
run_gp_bo <- function(X_init, y_init, budget, n_cand, kappa, eps = 1e-10, dup_tol = 1e-10) {
  X_eval <- as.matrix(X_init); y_eval <- as.numeric(y_init); d <- ncol(X_eval)
  best_so_far <- numeric(budget + 1); best_so_far[1] <- min(y_eval)
  
  for (t in 1:budget) {
    fit <- GP_fit(X = X_eval, Y = y_eval)
    X_cand <- maximinLHS(n_cand, d)
    pred <- predict(fit, xnew = X_cand)
    
    mu <- as.numeric(pred$Y_hat)
    sd_gp <- sqrt(pmax(as.numeric(pred$MSE), 0) + eps)
    
    # Acquisition: LCB
    lcb <- mu - kappa * sd_gp
    ord <- order(lcb)
    pick <- NA_integer_
    for (idx in ord) {
      if (!is_duplicate(X_cand[idx, ], X_eval, tol = dup_tol)) {
        pick <- idx; break
      }
    }
    if (is.na(pick)) pick <- ord[1]
    
    x_next <- X_cand[pick, , drop = FALSE]; y_next <- f(x_next)
    X_eval <- rbind(X_eval, x_next); y_eval <- c(y_eval, y_next)
    best_so_far[t + 1] <- min(y_eval)
  }
  list(X_eval = X_eval, y_eval = y_eval, best_so_far = best_so_far)
}
```

# Execution & Results

```{r run-experiment, message=FALSE, results='hide'}
res_bass <- run_bass_bo(X_init, y_init, budget, n_cand, kappa, dup_tol)
res_gp   <- run_gp_bo(X_init, y_init, budget, n_cand, kappa, eps, dup_tol)
res_rand <- run_random_search(X_init, y_init, budget, dup_tol)
```

## Convergence Comparison

```{r convergence-plot, fig.width=10, fig.height=5}
conv_df <- bind_rows(
  tibble(iter = 0:budget, best = res_bass$best_so_far, method = "BASS-BO"),
  tibble(iter = 0:budget, best = res_gp$best_so_far,   method = "GP-BO"),
  tibble(iter = 0:budget, best = res_rand$best_so_far, method = "Random Search")
)

ggplot(conv_df, aes(x = iter, y = best, color = method)) +
  geom_line(linewidth = 1) +
  labs(title = sprintf("Multi-Dimensional Convergence (d=%d)", d),
       x = "Iteration", y = "Best Objective Found") +
  theme_minimal()
```

























